Echoes from the scattering of wavepackets on wormholes

### 2018-12-15 paper.tex
no message

\section{Scattering of scalar wavepackets}\label{sec:scalar}
In this section, we solve the scattering of a Gaussian wavepacket by a spherically symmetric wormhole. To do so, we will first review the dispersion by the centrifugal barrier of a spherically symmetric black hole in order to find the properties of the potential cavity.
\subsection{Scattering by a Schwarzschild black hole}
The dispersion of scalar waves by a Schwarzschild black hole has been thoroughly studied \citep{doi:10.1063/1.522949, Sanchez:1976xm, Sanchez:1977si, Sanchez:1977vz}, it is crucial for our purposes to provide the full solution since it contains all the information of the potential barriers constituting the effective potential in the case of a wormhole. The dynamics of the scattering problem is found by solving the equation of motion for scalar field
\begin{equation}
\Box\Phi=0,\label{eq:scalar_eq_mov} 
\end{equation}
where $\Box\equiv g^{\alpha\beta}\nabla_{\alpha}\nabla_{\beta}$ is the standard d'Alembertian in a curved background. Here $g_{\alpha\beta}$ is the metric tensor in a spherically symmetric Schwarzschild-like static spacetime
\begin{equation}
g_{\alpha\beta}=-f(r)\delta^t_{\alpha}\delta^t_{\beta}+\frac{1}{f(r)}\delta^r_{\alpha}\delta^r_{\beta}+r^2\left(\delta^{\theta}_{\alpha}\delta^{\theta}_{\beta}+\sin^2\theta\delta^{\phi}_{\alpha}\delta^{\phi}_{\beta}\right).
\label{eq:Schwarzschild}
\end{equation}
\begin{figure}[t]
\centering
\includegraphics[width=.45\textwidth]{figures/potential_scalar_bh.pdf}
\caption{\label{fig:Potential_BH} Effective potential for the spherical modes $\mathcal{U}^{\mathrm{bh}}_{20}(x,t)$ scattered by a Schwarzschild black hole, growing with $\ell$. The wall acts as a barrier transparent to certain frequencies above a transmissivity threshold and reflective for lower frequencies.}
\end{figure}
It is convenient to introduce the tortoise coordinate $x$:
\begin{equation}
x\equiv\displaystyle{\int_{r_0}^r\frac{dr}{f(r)}},
\label{eq:tortoise}
\end{equation} 
In the case of the Schwarzschild metric $f(r)=1-r_g/r$ the last expression yields
\begin{equation}
x=r-r_0+r_g\ln\left(\frac{r-r_g}{r_0-r_g}\right),
\label{eq:tortoise2}
\end{equation} 
for $r_g<r<+\infty$ and $r_0>r_g$, where $r_g=2M$ is the usual Schwarzschild radius. By direct evaluation, we see that $r=r_0$ corresponds to $x=0$, the horizon $r=2M$ maps into $x\rightarrow-\infty$ and $r\rightarrow+\infty$ is $x\rightarrow+\infty$. In our numerical routine, we invert \eqref{eq:tortoise2} to get $r\equiv r(x)$ (see the appendix A, subsection 6 in \citep{Frolov:2017asg} for more details). In tortoise coordinates, we can decompose the scalar field in spherical harmonics 
\begin{equation}
\Phi(x,t)=\frac{1}{r(x)}\sum_{\ell,m=0}\mathcal{U}^{\mathrm{bh}}_{\ell m}(x,t)Y_{\ell m}(\theta,\phi),
\label{eq:ylm_decomp}
\end{equation}
in that way we can rewrite \eqref{eq:scalar_eq_mov} as
\begin{equation}
\left[-\partial_t^2+\partial_x^2-V_{\mathrm{scalar}}(x)\right]\mathcal{U}^{\mathrm{bh}}_{\ell m}(x,t) = 0,
\label{eq:wave_scalar}
\end{equation}
and the effective potential $V_{\mathrm{scalar}}(x)$ is given by
\begin{equation}
V_{\mathrm{scalar}}(x) = \left(1-\frac{r_g}{r(x)}\right)\left[\frac{\ell(\ell+1)}{r(x)^2}+\frac{r_g}{r(x)^3}\right].
\end{equation}
After rearranging the variables, the equation of motion of the spherical modes is now written in its traditional linear waveform. In Fig.~\ref{fig:Potential_BH}, we observe the growth of the potential barrier with the quantum number $\ell$. The potential wall does not vanish for the monopole ($\ell=0$) due to the extra term proportional to $r^{-3}$ appearing after the coordinate change, which replaces the radial damping in the original Schwarzschild coordinates $(t,r)$. Such a term becomes subdominant for all $\ell\geq 1$. Intuitively, it is reasonable to expect that the modes with frequency above a given threshold can cross the barrier, while reflecting the lower frequency modes.

@@

It is now possible to evaluate the reflection and transmission coefficients of the potential wall depicted in Fig.~\ref{fig:Potential_BH}. To do so, we compute the one dimensional Fourier transform of the incident $\tilde{\mathcal{U}}_{20}^{\mathrm{inc}}(\omega)=\mathcal{F}[\mathcal{U}^{\mathrm{bh}}_{20}(0,x)]$, reflected $\tilde{\mathcal{U}}_{20}^{\mathrm{ref}}(\omega)=\mathcal{F}[\mathcal{U}^{\mathrm{bh}}_{20}(t,+\infty)]$ and transmitted $\tilde{\mathcal{U}}_{20}^{\mathrm{trans}}(\omega)=\mathcal{F}[\mathcal{U}^{\mathrm{bh}}_{20}(t,-\infty)]$ from the solved scattering modes in order to define
\begin{equation}
R(\omega)\equiv \frac{||\tilde{\mathcal{U}}_{20}^{\mathrm{ref}}(\omega)||}{||\tilde{\mathcal{U}}_{20}^{\mathrm{inc}}(\omega)||}~,~T(\omega)\equiv \frac{||\tilde{\mathcal{U}}_{20}^{\mathrm{trans}}(\omega)||}{||\tilde{\mathcal{U}}_{20}^{\mathrm{inc}}(\omega)||}
\label{eq:ref_and_trans}
\end{equation}
\begin{figure}[t]
\centering
\includegraphics[width=.45\textwidth]{figures/potential_scalar.pdf}
\caption{\label{fig:potential_wh} Effective potential cavity for the wormhole, we observe the growth of the barriers with the quantum number $\ell$.}
\end{figure}

@@

\subsection{Scattering by a traversable wormhole} 
In this section, we study the dispersion of Gaussian wavepackets by a traversable wormhole, formed by the junction at $r_0>r_g$ of two Schwarzschild black hole solutions with equal mass. There is a discontinuity in $G_{\alpha\beta}$ such that at $r=r_0=20.0r_g$, any contracting congruence of geodesics in one side of the throat starts to expand in order to reach the other side, violating the weak energy condition.
The dynamics of the scattering problem is still given by the solutions of \eqref{eq:scalar_eq_mov} following the same decomposition in spherical modes as in \eqref{eq:ylm_decomp}. Hence, the waveform of the equation of motion for the spherical modes is given by
\begin{equation}
\left[-\partial_t^2+\partial_x^2-V^{\mathrm{wh}}_{\mathrm{scalar}}(x)\right]\mathcal{U}_{\ell m}(x,t) = 0,
\label{eq:wave_scalar_wh}
\end{equation}
and the effective potential $V^{\mathrm{wh}}_{\mathrm{scalar}}(x)$ yields
\begin{equation}
V^{\mathrm{wh}}_{\mathrm{scalar}}\left(x\right) = \left(1-\frac{r_g}{r\left(|x|\right)}\right)\left[\frac{\ell(\ell+1)}{r\left(|x|\right)^2}+\frac{r_g}{r\left(|x|\right)^3}\right].
\end{equation}   
\begin{figure}[t!]
\centering
\includegraphics[width=.3\textwidth]{figures/Geom_optics.pdf}
\caption{\label{fig:Geom_optics} Schematic reconstruction of the outgoing solution by a sequence of reflections and transmissions inside the potential cavity.}
\end{figure}
which is plotted in Fig.~\ref{fig:potential_wh} and coincides with the shape of the potential calculated in \citep{Cardoso:2016oxy}. Strictly speaking, we refer to $r\left(|x|\right)$ as the same inverse of the function mentioned in \eqref{eq:tortoise2} now evaluated at $|x|-r_0$. As we can see in Fig.~\ref{fig:potential_wh}, the new effective potential is merely a reflection of the potential barrier in Fig.~\ref{fig:Potential_BH} about the ordinate axis; thus, it is sensible to identify this system as a potential cavity built from two potential barriers with the reflection and transmission coefficients depicted in Fig.~\ref{fig:RT_scalar}. Furthermore, let us assume that an arbitrary incident pulse $\Phi_0$ propagates towards the cavity, it is, therefore, reasonable to approximate the spectrum of the asymptotic solution by a simple geometrical series of reflections and transmissions inside the cavity acting on the incident pulse, as shown in Fig.~\ref{fig:Geom_optics}. Following this scheme, the asymptotic solution can be approximated by
\begin{equation}
\Phi^{\mathrm{wh}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^{2i+1}\circ T\right]\circ\Phi_0(\omega).
\label{eq:reconst}
\end{equation}
It is not difficult to see that, in the hypothetic case of a perfectly reflective wall replacing the left potential barrier in Figs.~\ref{fig:potential_wh} and \ref{fig:Geom_optics}, the reconstructed pulse is instead given by 
\begin{equation}
\Phi^{\mathrm{f}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^i\circ T\right]\circ\Phi_0(\omega),
\label{eq:reconst_firewall}
\end{equation}
corresponding to the case of a firewall. We will not cover the features of the firewall solution in what remains of this paper.
\section{Scattering of tensor wavepackets}\label{sec:tensor} 
\section{Conclusions}\label{sec:conclusions}

>>>>

\section{Scattering of scalar wavepackets}\label{sec:scalar}
In this section, we solve the scattering of a Gaussian wavepacket by a spherically symmetric wormhole. To do so, we will first review the dispersion by the centrifugal barrier of a spherically symmetric black hole in order to find the properties of the potential cavity.
\subsection{Scattering by a Schwarzschild black hole}\label{subsec:bh} 
The dispersion of scalar waves by a Schwarzschild black hole has been thoroughly studied \citep{doi:10.1063/1.522949, Sanchez:1976xm, Sanchez:1977si, Sanchez:1977vz}, it is crucial for our purposes to provide the full solution since it contains all the information of the potential barriers constituting the effective potential in the case of a wormhole. The dynamics of the scattering problem is found by solving the equation of motion for scalar field
\begin{equation}
\Box\Phi=0,\label{eq:scalar_eq_mov} 
\end{equation}
where $\Box\equiv g^{\alpha\beta}\nabla_{\alpha}\nabla_{\beta}$ is the standard d'Alembertian in a curved background. Here $g_{\alpha\beta}$ is the metric tensor in a spherically symmetric Schwarzschild-like static spacetime
\begin{equation}
g_{\alpha\beta}=-f(r)\delta^t_{\alpha}\delta^t_{\beta}+\frac{1}{f(r)}\delta^r_{\alpha}\delta^r_{\beta}+r^2\left(\delta^{\theta}_{\alpha}\delta^{\theta}_{\beta}+\sin^2\theta\delta^{\phi}_{\alpha}\delta^{\phi}_{\beta}\right).
\label{eq:Schwarzschild}
\end{equation}
\begin{figure}[t]
\centering
\includegraphics[width=.45\textwidth]{figures/potential_scalar_bh.pdf}
\caption{\label{fig:Potential_BH} Effective potential for the spherical modes $\mathcal{U}^{\mathrm{bh}}_{20}(x,t)$ scattered by a Schwarzschild black hole, growing with $\ell$. The wall acts as a barrier transparent to certain frequencies above a transmissivity threshold and reflective for lower frequencies.}
\end{figure}
It is convenient to introduce the tortoise coordinate $x$:
\begin{equation}
x\equiv\displaystyle{\int_{r_0}^r\frac{dr}{f(r)}},
\label{eq:tortoise}
\end{equation} 
In the case of the Schwarzschild metric $f(r)=1-r_g/r$ the last expression yields
\begin{equation}
x=r-r_0+r_g\ln\left(\frac{r-r_g}{r_0-r_g}\right),
\label{eq:tortoise2}
\end{equation} 
for $r_g<r<+\infty$ and $r_0>r_g$, where $r_g=2M$ is the usual Schwarzschild radius. By direct evaluation, we see that $r=r_0$ corresponds to $x=0$, the horizon $r=2M$ maps into $x\rightarrow-\infty$ and $r\rightarrow+\infty$ is $x\rightarrow+\infty$. In our numerical routine, we invert \eqref{eq:tortoise2} to get $r\equiv r(x)$ (see the appendix A, subsection 6 in \citep{Frolov:2017asg} for more details). In tortoise coordinates, we can decompose the scalar field in spherical harmonics 
\begin{equation}
\Phi(x,t)=\frac{1}{r(x)}\sum_{\ell,m=0}\mathcal{U}^{\mathrm{bh}}_{\ell m}(x,t)Y_{\ell m}(\theta,\phi),
\label{eq:ylm_decomp}
\end{equation}
in that way we can rewrite \eqref{eq:scalar_eq_mov} as
\begin{equation}
\left[-\partial_t^2+\partial_x^2-V_{\mathrm{scalar}}(x)\right]\mathcal{U}^{\mathrm{bh}}_{\ell m}(x,t) = 0,
\label{eq:wave_scalar}
\end{equation}
and the effective potential $V_{\mathrm{scalar}}(x)$ is given by
\begin{equation}
V_{\mathrm{scalar}}(x) = \left(1-\frac{r_g}{r(x)}\right)\left[\frac{\ell(\ell+1)}{r(x)^2}+\frac{r_g}{r(x)^3}\right].
\end{equation}
After rearranging the variables, the equation of motion of the spherical modes is now written in its traditional linear waveform. In Fig.~\ref{fig:Potential_BH}, we observe the growth of the potential barrier with the quantum number $\ell$. The potential wall does not vanish for the monopole ($\ell=0$) due to the extra term proportional to $r^{-3}$ appearing after the coordinate change, which replaces the radial damping in the original Schwarzschild coordinates $(t,r)$. Such a term becomes subdominant for all $\ell\geq 1$. Intuitively, it is reasonable to expect that the modes with frequency above a given threshold can cross the barrier, while reflecting the lower frequency modes.

@@

It is now possible to evaluate the reflection and transmission coefficients of the potential wall depicted in Fig.~\ref{fig:Potential_BH}. To do so, we compute the one dimensional Fourier transform of the incident $\tilde{\mathcal{U}}_{20}^{\mathrm{inc}}(\omega)=\mathcal{F}[\mathcal{U}^{\mathrm{bh}}_{20}(0,x)]$, reflected $\tilde{\mathcal{U}}_{20}^{\mathrm{ref}}(\omega)=\mathcal{F}[\mathcal{U}^{\mathrm{bh}}_{20}(t,+\infty)]$ and transmitted $\tilde{\mathcal{U}}_{20}^{\mathrm{trans}}(\omega)=\mathcal{F}[\mathcal{U}^{\mathrm{bh}}_{20}(t,-\infty)]$ from the solved scattering modes in order to define
\begin{equation}
R(\omega)\equiv \frac{||\tilde{\mathcal{U}}_{20}^{\mathrm{ref}}(\omega)||}{||\tilde{\mathcal{U}}_{20}^{\mathrm{inc}}(\omega)||}~,~T(\omega)\equiv \frac{||\tilde{\mathcal{U}}_{20}^{\mathrm{trans}}(\omega)||}{||\tilde{\mathcal{U}}_{20}^{\mathrm{inc}}(\omega)||}
\label{eq:ref_and_trans}
\end{equation}
\begin{figure}[t]
\centering
\includegraphics[width=.4\textwidth]{figures/potential_scalar.pdf}
\caption{\label{fig:potential_wh} Effective potential cavity for the wormhole, we observe the growth of the barriers with the quantum number $\ell$.}
\end{figure}

@@

\subsection{Scattering by a traversable wormhole}\label{subsec:wh} 
In this section, we study the dispersion of Gaussian wavepackets by a traversable wormhole, formed by the junction at $r_0>r_g$ of two Schwarzschild black hole solutions with equal mass. There is a discontinuity in $G_{\alpha\beta}$ such that at $r=r_0=20.0r_g$, any contracting congruence of geodesics in one side of the throat starts to expand in order to reach the other side, violating the weak energy condition.
The dynamics of the scattering problem is still given by the solutions of \eqref{eq:scalar_eq_mov} following the same decomposition in spherical modes as in \eqref{eq:ylm_decomp}. Hence, the waveform of the equation of motion for the spherical modes is given by
\begin{equation}
\left[-\partial_t^2+\partial_x^2-V^{\mathrm{wh}}_{\mathrm{scalar}}(x)\right]\mathcal{U}_{\ell m}(x,t) = 0,
\label{eq:wave_scalar_wh}
\end{equation}
and the effective potential $V^{\mathrm{wh}}_{\mathrm{scalar}}(x)$ yields
\begin{equation}
V^{\mathrm{wh}}_{\mathrm{scalar}}\left(x\right) = \left(1-\frac{r_g}{r\left(|x|\right)}\right)\left[\frac{\ell(\ell+1)}{r\left(|x|\right)^2}+\frac{r_g}{r\left(|x|\right)^3}\right].
\end{equation}   
\begin{figure}[t!]
\centering
\includegraphics[width=.3\textwidth]{figures/Geom_optics.pdf}
\caption{\label{fig:Geom_optics} Schematic reconstruction of the outgoing solution by a sequence of reflections and transmissions inside the potential cavity.}
\end{figure}
which is plotted in Fig.~\ref{fig:potential_wh} and coincides with the shape of the potential calculated in \citep{Cardoso:2016oxy}. Strictly speaking, we refer to $r\left(|x|\right)$ as the same inverse of the function mentioned in \eqref{eq:tortoise2} now evaluated at $|x|-r_0$. As we can see in Fig.~\ref{fig:potential_wh}, the new effective potential is merely a reflection of the potential barrier in Fig.~\ref{fig:Potential_BH} about the ordinate axis; thus, it is sensible to identify this system as a potential cavity built from two potential barriers with the reflection and transmission coefficients depicted in Fig.~\ref{fig:RT_scalar}. Furthermore, let us assume that an arbitrary incident pulse $\Phi_0$ propagates towards the cavity, it is, therefore, reasonable to approximate the spectrum of the asymptotic solution by a simple geometrical series of reflections and transmissions inside the cavity acting on the incident pulse, as shown in Fig.~\ref{fig:Geom_optics}. Following this scheme, the asymptotic solution can be approximated by
\begin{equation}
\Phi^{\mathrm{wh}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^{2i+1}\circ T\right]\circ\Phi_0(\omega).
\label{eq:reconst}
\end{equation}
It is not difficult to see that, in the hypothetic case of a perfectly reflective wall replacing the left potential barrier in Figs.~\ref{fig:potential_wh} and \ref{fig:Geom_optics}, the reconstructed pulse is instead given by 
\begin{equation}
\Phi^{\mathrm{f}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^i\circ T\right]\circ\Phi_0(\omega),
\label{eq:reconst_firewall}
\end{equation}
corresponding to the case of a firewall. We will not cover the features of the firewall solution in what remains of this paper.

\begin{figure*}
\centering
\subfigure{
\includegraphics[width=.4\textwidth]{figures/wh_l_2.pdf}} \,
\subfigure{
\includegraphics[width=.4\textwidth]{figures/wh_flux_l_2.pdf}}
\caption{\label{fig:sol_n_flux_wh} Left panel: Evolution of the quadrupole mode $\mathcal{U}_{20}$ for the effective potential in Fig.~\ref{fig:potential_wh}. We can notice the sequence of reflections and transmissions inside the cavity is very similar to the scheme depicted in Fig.~\ref{fig:Geom_optics}. Right panel: Evolution of the radial flux $\mathcal{J}_{20}\equiv \mathcal{U}_{20,x} \mathcal{U}_{20,t}$, the only incident source comes from the collision of the pulse, which dissipates very slowly to the exterior every time the internal reflections hit the walls of the potential cavity. QNM are sourced by this process.} 
\end{figure*}

We now solve the scattering problem exactly for the quadrupole mode $\mathcal{U}_{20}(t,x)$. Using the same initial conditions as in \eqref{eq:scalar_init_cond} for $\sigma=0.9185r_g$ as the width of the incident Gaussian pulse, we find the time-dependent solution of the quadrupole mode $\mathcal{U}_{20}$ (left panel) and its radial flux $\mathcal{J}_{20}\equiv \mathcal{U}_{20,x} \mathcal{U}_{20,t}$ (right panel) in Fig.~\ref{fig:sol_n_flux_wh}. It is interesting to notice in the evolution plot (on the left) that the signal forms an interference pattern at very late times, showing that this process might fill the cavity. In addition to this, even when the amplitude of the modes decreases after each collision against any of the potential walls, the spherical modes propagate for longer time throughout the spheres of maximal potential.

The amplitude of the outgoing signals depends on the width of the incident Gaussian wavepacket, this is visible in Fig.~\ref{fig:asympt_two_sigmas} where we plot the asymptotic solutions for two different ingoing wavepackets: one with $\sigma=0.6495r_g$ in the left panel and a second one with $\sigma=5.196r_g$ in the right. After subtracting the outgoing solution for a black hole, the presence of a train of wavelets, widely known as echoes, is very clear. The signal is not a straight line after the transient, as we can observe in the right panel of this figure. Therefore, subtracting the black hole outgoing pulse (i.e., the case in which there is only one potential wall) turns to be a convenient way to clean the signal from back reflections due to the ``tails'' of the potential barrier, especially since in this case the echoes are four orders of magnitude smaller than the transient.

As shown in subsection \ref{subsec:bh}, the values of reflectivity and transmissivity determine the frequencies going through the walls of the potential cavity: most of the power of an incident pulse with large $\sigma$ is in the low frequency domain, and therefore it will be reflected. Moreover, the cavity is transparent to high frequency signals, which are dominant in the pulses with small $\sigma$. 
\begin{figure*}
\centering
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Scalar_echo_w_06495.pdf}} \,
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Scalar_echo.pdf}}
\caption{\label{fig:asympt_two_sigmas} Left panel: Asymptotic evolution of the quadrupole mode $\mathcal{U}_{20}$ for $\sigma=0.6495r_g$. Echoes are plotted in the upper corner of the plot, showing them along with their Hilbert envelope (in green) and maximum amplitudes (in red). Right panel: Asymptotic evolution of the quadrupole mode $\mathcal{U}_{20}$ for $\sigma=5.196r_g$. In contrast with the left panel, echoes are four orders of magnitude smaller than the transient.} 
\end{figure*}
In either of these extremal scenarios, QNMs cannot be sourced by internal reflections and thus, the amplitude of the echoes is not generically large. Now we proceed to reconstruct the asymptotic spectrum using the reflectivity and transmissivity defined in \eqref{eq:ref_and_trans} and the compare it with the Fourier transformed asymptotic solutions of \eqref{eq:wave_scalar_wh}.

\section{Scattering of tensor wavepackets}\label{sec:tensor} 
\section{Conclusions}\label{sec:conclusions}
